Automatic Identification and Annotation of Parallel Tasks in Structured Programs

### 2017-10-31 docs/ACM/paper.tex

\title[Automatic Identification and Annotation of Parallel Tasks in Structured
Programs]
{Automatic Identification and Annotation of Parallel Tasks in Programs}

@@

Annotation systems have risen to a place of prominence as a simple and
effective way to write parallel programs.
Examples of such systems include OpenMP~\cite{JaegerCP15},
OpenACC~\cite{OpenACC20}, OpenHMPP~\cite{Andion14}, OpenMPC~\cite{Lee10} and OpenSs~\cite{MeenderinckJ11}.
Annotations work as a meta-language: they let developers grant parallel
semantics to syntax originally written to execute sequentially.
The compiler-related literature contains a generous load of success stories
describing the use of annotation systems to parallelize programs.
Combined with modern accelerators, they have led to substantial performance
gains~\cite{Bertolli14,Mendonca17,Poesia17,Reyes12,Wienke12}.
Yet, even if convenient, the use of annotations is not straightforward.

Annotations such as OpenMP or OpenACC still require developers to
worry about typical hassles of the parallel world, such as race conditions,
or deadlocks.
Moreover, because they are used in tandem with imperative programming languages,
such troubles are only made worse by pointer aliasing.
There exist tools that insert automatic annotations in
programs~\cite{gleison2016,Pingali11,wanggenerating}.
All these technologies explore data-parallelism -- the possibility of running
the same computation independently on different data.
Task parallelism remains still uncharted land, in what concerns automatic
annotation.
Such fact is unfortunate, because much of the power of current annotation
systems lays on their ability to create tasks \ed{We need to find references!}.
This ability brings said systems closer to irregular programs such as those
that implement graphs and worklist algorithms \`{a} la
Pingali~\cite{Ben-Nun17,Kulkarni11,Pingali11}.
We address this omission in this work.

This paper describes \Taskminer, the first compiler that mines task parallelism
from programs.
To fulfill this goal \Taskminer{} solves three challenges.
First, it find program regions, e.g., loops or functions, that can be
effectively mapped onto tasks.
Second, it determines symbolic bounds to all the memory blocks accesses within
those regions.
Third, it maps such information back into source code, producing annotations
that programmers can read.
The goal of this paper is to describe the design and the implementation of these
three technologies, and to demonstrate their benefit.

Such benefit is performance and readability.
\Taskminer{} receives as input a C program, and produces, as output, a C
program annotated with OpenMP directives that identify tasks.
We are currently able to annotate non-trivial programs, involving every
sort of composite type available in the C language, e.g., arrays, structs,
unions and pointers to these aggregates.
Some of these programs, such as those taken from the
Kastor~\cite{Virouleau14} or Bots~\cite{Duran09}
benchmark suites, are large and complex; thus, their manual annotation is a
time consuming and error prone task.
Yet, our automatically annotated programs not only approximate the execution
times of the parallel versions of those benchmarks, but are much faster than
their sequential -- unannotated -- versions.

@@

%EXAMPLE - BFS

@@

>>>>

\title[Automatic Identification and Annotation of Parallel Tasks in Structured
Programs]
{Automatic Identification and Annotation of Parallel Tasks in Structured Programs}

@@

Annotation systems have risen to a place of prominence as a simple and
effective way to write parallel programs.
Examples of such systems include OpenMP~\cite{JaegerCP15},
OpenACC~\cite{OpenACC20}, OpenHMPP~\cite{Andion14}, OpenMPC~\cite{Lee10} and OpenSs~\cite{MeenderinckJ11}.
Annotations work as a meta-language: they let developers grant parallel
semantics to syntax originally written to execute sequentially.
The compiler-related literature contains a generous load of success stories
describing the use of annotation systems to parallelize programs.
Combined with modern accelerators, they have led to substantial performance
gains~\cite{Bertolli14,Mendonca17,Poesia17,Reyes12,Wienke12}.
Yet, even if convenient, the use of annotations is not straightforward, and
lacks supporting tools.

Annotations such as OpenMP or OpenACC still require developers to
worry about typical hassles of the parallel world, such as race conditions
or deadlocks.
Moreover, because they are used in tandem with imperative programming languages,
such troubles are only made worse by pointer aliasing.
There exist tools that insert automatic annotations in
programs~\cite{gleison2016,Pingali11,wanggenerating}.
All these technologies explore data-parallelism -- the possibility of running
the same computation independently on different data.
Task parallelism remains still uncharted land, in what concerns automatic
annotation.
Such fact is unfortunate, because much of the power of current annotation
systems lays on their ability to create tasks \ed{We need to find references!}.
This ability brings said systems closer to irregular programs such as those
that implement graphs and worklist algorithms \`{a} la
Pingali~\cite{Ben-Nun17,Kulkarni11,Pingali11}.
The goal of this work is to address this omission.

This paper describes \Taskminer, the first compiler that mines task parallelism
from programs.
To fulfill this goal \Taskminer{} solves three challenges.
First, it find program regions, e.g., loops or functions, that can be
effectively mapped onto tasks (Section~\ref{sub:region}).
Second, it determines symbolic bounds to all the memory blocks accessed within
those regions (Section~\ref{sub:symb}).
Third, it maps such information back into source code, producing annotations
that programmers can read (Section~\ref{sub:ir}).
The goal of this paper is to describe the design and the implementation of these
three technologies, and to demonstrate their benefit.

Such benefit is performance and readability.
\Taskminer{} receives as input a C program, and produces, as output, a C
program annotated with OpenMP directives that identify tasks.
We are currently able to annotate non-trivial programs, involving every
sort of composite type available in the C language, e.g., arrays, structs,
unions and pointers to these aggregates.
Some of these programs, such as those taken from the
Kastor~\cite{Virouleau14} or Bots~\cite{Duran09}
benchmark suites, are large and complex; thus, their manual annotation is a
time consuming and error prone task.
Yet, our automatically annotated programs not only approximate the execution
times of the parallel versions of those benchmarks, but are much faster than
their sequential --unannotated-- versions, as we report in Section~\ref{sec:eval}.

@@

@@

\subsection{Mapping Program Regions to Tasks}
\label{sub:region}

\subsection{Find Symbolic Bounds of Arrays}
\label{sub:symb}

\subsection{From IR to Source Code}
\label{sub:ir}
